---
title: "Access UKB data on Rosalind"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to access UKB phenotype data on KCL Rosalind HPC cluster}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  echo = TRUE,
  message = FALSE
)

requireNamespace("devtools", quietly = TRUE)
requireNamespace("stringr", quietly = TRUE)

devtools::load_all(".")
```


***

**Important. This document is only useful for UKB-approved KCL reasearchers and their collaborators, with an account on the Rosalind HPC cluster.**

The ukbkings package works with a Rosalind UKB project directory that has been setup to use the package. See vignette("Setup UKB project on Rosalind") or read the documentation under Articles.

***

<span style="color:dodgerblue;">**Contents:**</span>  
[0. tldr]  
[1. Preamble]  
[2. Installation]  
[3. Field subset file]  
[4. Read and write data]  
[5. Categorical codes]  
[6. GP data]  
[7. COVID-19 data]  
[8. HES in-patient data]  
[9. Death data]

***

### 0. tldr

Install the package
```{R}
devtools::install_github("kenhanscombe/ukbkings", dependencies = TRUE, force = TRUE)
```

Write a serialised R dataframe to file for your required fields.
```{R}
bio_phen(
 project_dir = "<path_to_project_directory>",
 field = "<path_to_required_fields_file>",  # one per line, no header
 out = "<stem_of_output_file>"  # e.g. "data/ukb" writes "data/ukb.rds"
)
```

***Note**. `bio_phen` reads a withdrawal file from the project
directory and replaces phenotype values for to-be-excluded samples
with `NA`.*

<br>

Read the generated dataset into R.
```{R}
df <- readRDS("data/ukb.rds")
```

<br>

### 1. Preamble

On Rosalind, change to your UKB user directory. If you do not have a
UKB user directory, create one first at /scratch/groups/ukbiobank/Edinburgh_Data/usr/*\<username\>*. Start an interactive cluster session with sufficient memory to read the UKB data (I think you can get away with less, but I've not worked that out yet). Load the default cluster version of R (`module avail 2&>1 | grep R` currently `apps/R/3.6.0`) then start R.
```{bash}
# Rosalind shell
cd /scratch/groups/ukbiobank/Edinburgh_Data/usr/<username>
srun -p shared,brc --mem=30G --pty /bin/bash

module load apps/R/3.6.0
R
```

<br>

### 2. Installation

***Note**. All code blocks below are R, unless otherwise specified.*

<br>

Install from Github
```{R}
devtools::install_github("kenhanscombe/ukbkings", dependencies = TRUE, force = TRUE)
```

Load libraries. (I like to use tidyverse but it is not necessary, base R is fine)
```{R}
library(ukbkings)
library(tidyverse)
```

Check help (press 'q' to exit).
```{R}
?ukbkings
```

***Note**. All included data and functionality is described in the
package index and also described on the
[ukbkings webpage](https://kenhanscombe.github.io/ukbkings/) under the **Reference** tab.*

<br>

Check help on specific function, e.g.,
```{R}
?bio_phen
```

Point to the project directory.
```{R}
project_dir <- "/scratch/datasets/ukbiobank/ukb<project_id>_<ownername>"
```

<br>

### 3. Field subset file

You need a file with required fields, one per line, no header.

<br>

Read the project field-to-name "field finder" file, inspect the variable metadata, and display the number of baskets included.
```{R}
f <- bio_field(project_dir)

head(f)
glimpse(f)

f %>%
distinct(basket)
```

Search for variables required and add their field codes to a file, one per line, no header. You can page through the file.
```{R}
f %>%
select(name) %>%
page(method = "print")
```

Or, search `name` column
```{R}
f %>%
select(field, name) %>%
filter(str_detect(name, "vegetables"))

f %>%
select(field, name) %>%
filter(str_detect(name, "ldl|triglycerides"))
```

Alternatively, search the [UKB showcase](http://biobank.ndph.ox.ac.uk/showcase/) for a variable of interest then filter on the `field` column in the field-to-name dataframe (useful if multiple instances required). For example, if you search for "cholesterol medication", the field stem you want is 6177
```{R}
f %>%
select(field, name) %>%
filter(str_detect(field, "6177"))
```

`bio_field_add` is a convenience function for creating the one per line required variables/fields file. By default the function appends fields. Create the field subset file in your UKB user directory /scratch/groups/ukbiobank/Edinburgh_Data/usr/*\<username\>*/
```{R}
f %>%
select(field, name) %>%
filter(str_detect(field, "6177")) %>%
bio_field_add("small_field_subset.txt")
```

Inspect the field selection file.
```{R}
system("cat small_field_subset.txt")
```

<br>

### 4. Read and write data

Read required fields and save as an rds file in your user directory. Argument `out` should be a path to your UKB user directory
```{r}
bio_phen(
 project_dir,
 field = "small_field_subset.txt",
 out = "small_phenotype_subset"
)
```

***Note**. Dates in the UKB data are recorded in a variety of formats, some of which are non-standard: "character string is not in a standard unambiguous format", e.g., 2009-01-12T11:28:56. All date variables have been left in character format for the user to convert as needed.*

<br>

Check the size of your file and read in your dataset
```{r}
system("ls -lh small_phenotype_subset.rds")
df <- readRDS("small_phenotype_subset.rds")
```

If required, rename columns from the default UKB field names to the descriptive names used in the field-to-name "field finder" `name` column.

```{R}
df <- bio_rename(df, f)
```

<br>

### 5. Categorical codes

Categorical field codings are included in the field finder
```{R}
f %>%
 filter(field %in% names(df)) %>%
 select(field, categorical_coding)
```

Retrieve numerical "Value" and and associated "Meaning" for each categorical code.
```{R}
cx <- bio_code(project_dir)
head(cx)
```

Look up a particular coding
```{R}
cx %>%
 filter(Coding == 502)
```

<br>

### 6. GP data

If your project has access to primary care data, retrieve the clinical, registrations, scripts datasets with `bio_gp`
```{R}
gp_clinical <- bio_gp(project_dir, "clinical")
gp_registrations <- bio_gp(project_dir, "registrations")
gp_scripts <- bio_gp(project_dir, "scripts")
```

<br>

For details see UKB documentation:

* [Category 3001](http://biobank.ndph.ox.ac.uk/showcase/label.cgi?id=3001)
* [Resource 591: Primary Care Linked Data](http://biobank.ndph.ox.ac.uk/showcase/refer.cgi?id=591)
* [Resource 592: Clinical coding classification systems and maps](http://biobank.ndph.ox.ac.uk/showcase/refer.cgi?id=592)

<br>

### 7. COVID-19 data

If your project has access to COVID-19 data, retrieve the results with `bio_covid`
```{R}
covid <- bio_covid(project_dir, "results")
```

Read in the particapant blood group data extracted from imputed genotype
```{r}
covid_misc <- bio_covid(project_dir, "misc")
```

Descriptive meanings of the numerical values used to encode the categorical variables can be retrieved with `data = codes`
```{R}
covid_codes <- bio_covid(project_dir, "codes")
```

<br>

For details see UKB documentation:

* [Data field 40100](http://biobank.ndph.ox.ac.uk/showcase/field.cgi?id=40100)  
* [COVID-19 test results data](http://biobank.ndph.ox.ac.uk/showcase/exinfo.cgi?src=COVID19_tests)

<br>

### 8. HES in-patient data

To retrieve detailed information on hospital episode statistic use `bio_hesin`
```{r}
hesin_critical <- bio_hesin(project_dir, "critical")
hesin_delivery <- bio_hesin(project_dir, "delivery")
hesin_diag <- bio_hesin(project_dir, "diag")
# etc.
```

For all available data see `?bio_hesin`

<br>

For details see UKB documentation:

* [Category 2000](https://biobank.ndph.ox.ac.uk/showcase/label.cgi?id=2000)
* [Category 2006](https://biobank.ndph.ox.ac.uk/showcase/label.cgi?id=2006)
* [Resource 138483: Hospital Episode Statistics](https://biobank.ndph.ox.ac.uk/showcase/refer.cgi?id=138483)
* [Resource 141140: HES Data Dictionary](https://biobank.ndph.ox.ac.uk/showcase/refer.cgi?id=141140)

<br>

### 9. Death data

For data on date and cause of death (ICD10) obtained through linkage to national death registries, use `bio_death`
```{r}
death_date <- bio_death(project_dir, "death")
death_cause <- bio_death(project_dir, "cause")
```

<br>

For details see UKB documentation:

* [Category 100093](https://biobank.ctsu.ox.ac.uk/crystal/label.cgi?id=100093)
* [Resource 134993: Health Outcomes Report](https://biobank.ctsu.ox.ac.uk/crystal/refer.cgi?id=134993)
