---
title: "How to access UKB data on Rosalind"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to access UKB phenotype data on KCL Rosalind HPC cluster}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  echo = TRUE,
  message = FALSE
)

requireNamespace("devtools", quietly = TRUE)
requireNamespace("stringr", quietly = TRUE)

devtools::load_all(".")
```


***

**Note. This document is only useful for UKB-approved KCL reasearchers and their collaborators, with an account on the Rosalind HPC cluster.**

***

### TL;DR

Install the package
```{R}
devtools::install_github("kenhanscombe/ukbkings", dependencies = TRUE, force = TRUE)
```

Write a serialised R dataframe to file for your required fields.
```{R}
bio_phen(
 project_dir = "<path_to_project_directory>",
 field = "<path_to_required_fields_file>",  # one per line, no header
 out = "<stem_of_output_file>"  # e.g. "data/ukb" writes "data/ukb.rds"
)
```

Read it into R, e.g.,
```{R}
df <- readRDS("data/ukb.rds")
```

**Note**. `bio_phen` reads all withdrawals from a project subdirectory (w*.csv withdrawal files received from the UKB by email), and replaces their phenotype values with `NA`.

<br>

### PREAMBLE

Log in to Rosalind.
```{bash}
# Local shell
ssh <username>@login3.rosalind.kcl.ac.uk
```

Change to your UKB user directory. If you do not have a UKB user directory, create one first at /scratch/groups/ukbiobank/Edinburgh_Data/usr/*\<username\>*. Start an interactive cluster session with sufficient memory to read the UKB data (I think you can get away with less, but I've not worked that out yet). Load the default cluster version of R (`module avail 2&>1 | grep R` currently `apps/R/3.6.0`) then start R.
```{bash}
# Rosalind shell
cd /scratch/groups/ukbiobank/Edinburgh_Data/usr/<username>
srun -p shared,brc --mem=30G --pty /bin/bash

module load apps/R/3.6.0
R
```

<br>

### INSTALLATION/ SETUP

***Note**. All code blocks below are R, unless otherwise specified.*

<br>

Install from Github
```{R}
devtools::install_github("kenhanscombe/ukbkings", dependencies = TRUE, force = TRUE)
```


***

***Note**. If this fails, try create an environment from the included yml file. Activate it and try the install again.*
```{bash}
# Rosalind shell
wget https://raw.githubusercontent.com/kenhanscombe/ukbkings/master/environment.yml

conda env create -f environment.yml
source activate ukb
R
```

***


Load libraries. (I like to use tidyverse but it is not necessary, base R is fine)
```{R}
library(ukbkings)
library(tidyverse)
```

Check help (press 'q' to exit).
```{R}
?ukbkings
```

List functions available with `ukbkings::` TAB TAB. All basic data collection functions start `bio_`. Check help on specific function, e.g.,
```{R}
?bio_phen
```


***Note**. All functionality is also described on the [ukbkings webpage](https://kenhanscombe.github.io/ukbkings/) under the **Reference** tab.*

<br>

Point to the project directory, e.g., ukb18177_glanville
```{R}
project_dir <- "/scratch/datasets/ukbiobank/ukb18177_glanville"
```

<br>

### FIELD SUBSET FILE

You need a file with required fields, one per line, no header.

<br>

Read the project field-to-name "field finder" file, inspect the variable metadata, and display the number of baskets included.
```{R}
f <- bio_field(project_dir)

head(f)
glimpse(f)

f %>%
distinct(basket)
```

Search for variables required and add their field codes to a file, one per line, no header. You can page through the file.
```{R}
f %>%
select(name) %>%
page(method = "print")
```

Or, search `name` column
```{R}
f %>%
select(field, name) %>%
filter(str_detect(name, "vegetables"))

f %>%
select(field, name) %>%
filter(str_detect(name, "ldl|triglycerides"))
```

Alternatively, search the [UKB showcase](http://biobank.ndph.ox.ac.uk/showcase/) for a variable of interest then filter on the `field` column in the field-to-name dataframe (useful if multiple instances required). For example, if you search for "cholesterol medication", the field stem you want is 6177
```{R}
f %>%
select(field, name) %>%
filter(str_detect(field, "6177"))
```

`bio_field_add` is a convenience function for creating the one per line required variables/fields file. By default the function appends fields. Create the field subset file in your UKB user directory /scratch/groups/ukbiobank/Edinburgh_Data/usr/*\<username\>*/
```{R}
f %>%
select(field, name) %>%
filter(str_detect(field, "6177")) %>%
bio_field_add("small_field_subset.txt")
```

Inspect the field selection file.
```{R}
system("cat small_field_subset.txt")
```

<br>

### WRITE/ READ DATA

Read required fields and save as an rds file in your user directory. Argument `out` should be a path to your UKB user directory
```{r}
bio_phen(
 project_dir,
 field = "small_field_subset.txt",
 out = "small_phenotype_subset"
)
```

***Note**. Dates in the UKB data are recorded in a variety of formats, some of which are non-standard: "character string is not in a standard unambiguous format", e.g., 2009-01-12T11:28:56. All date variables have been left in character format for the user to convert as needed.*

<br>

Check the size of your file and read in your dataset
```{r}
system("ls -lh small_phenotype_subset.rds")
df <- readRDS("small_phenotype_subset.rds")
```

<br>

### CATEGORICAL CODES

Categorical field codings are included in the field finder
```{R}
f %>%
 filter(field %in% names(df)) %>%
 select(field, categorical_coding)
```

Retrieve numerical "Value" and and associated "Meaning" for each categorical code.
```{R}
cx <- bio_code(project_dir)
head(cx)
```

Look up a particular coding
```{R}
cx %>%
 filter(Coding == 502)
```

<br>

### GP DATA

If your project has access to primary care data, retrieve the clinical, registrations, scripts datasets with `bio_gp` e.g.
```{R}
px <- bio_gp(project_dir, "scripts")
head(px)
```
