---
title: "Using UKB primary care prescription data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using UKB primary care prescription data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    eval = FALSE,
    collapse = TRUE,
    comment = "#>"
)
```


UKB primary care presciption data are included in the
[Primary Care Linked data](https://biobank.ndph.ox.ac.uk/showcase/showcase/docs/primary_care_data.pdf),
accessed through the UKB Data Portal. The data can be
explored online with SQL queries, the results of which can be
downloaded. Alternatively, the full table can be downloaded.

For each KCL UKB application with access to primary care prescription
data, we have downloaded the full table and converted it to an R
disk.frame for fast local queries with ukbkings `bio_record` and
`bio_record_map`. (See the article
[Access UKB data on Rosalind/CREATE](https://kenhanscombe.github.io/ukbkings/articles/kcl-ukb-access.html#record-level-data)
for a detailed description of the use of `bio_record` and `bio_record_map` to query all
record-level data available to a particular KCL UKB application)

<br>

## Query prescription data

`bio_record` returns a diskframe, an object which can be queried in
chunks without reading the data into R.

```{r}
library(ukbkings)
library(dplyr)
library(stringr)
library(readxl)
library(lubridate)

project_dir <- "<path_to_project_directory>"

diskf <- bio_record(project_dir, record = "gp_scripts")
head(diskf)
```

<span style="background-color:grey">
<pre>
       eid data_provider issue_date read_2 bnf_code dmd_code                                    drug_name quantity
1: XXXXXXX             X XX/XX/XXXX  bi33. 02050501     <NA>                      Lisinopril 10mg tablets   56.000
2: XXXXXXX             X XX/XX/XXXX  blb1. 02060200     <NA>                       Amlodipine 5mg tablets   56.000
3: XXXXXXX             X XX/XX/XXXX  bxd5. 02120400     <NA>                     Simvastatin 40mg tablets   56.000
4: XXXXXXX             X XX/XX/XXXX  l412. 12010151     <NA> Otosporin ear drops (GlaxoSmithKline UK Ltd)   10.000
5: XXXXXXX             X XX/XX/XXXX  bi33. 02050501     <NA>                      Lisinopril 10mg tablets   28.000
6: XXXXXXX             X XX/XX/XXXX  blb1. 02060200     <NA>                       Amlodipine 5mg tablets   28.000
</pre>
</span>

<br>

You can use dplyr verbs to `filter` rows, `select` columns, etc., and
finally `collect` to retrieve the resulting data into a local dataframe.
Below is an example of using a combination of `bio_record` and dplyr
verbs to retrieve antidepressant prescriptions.

<br>

The primary care prescription data include 3 prescription coding systems

| code | meaning |
|---|---|
| bnf_code | British National Formulary (BNF) is a reference book containing the standard list of medicines used in UK |
| dmd_code | Dictionary of Medicines and Devices (dm+d), the catalogue of things that can be prescribed in the NHS |
| read_2 | Read Codes (version 2 (v2)) are a coded thesaurus of clinical terms (used in the NHS since 1985). There is also a version 3 (CTV3 or v3)  |


To retrieve prescriptions for a particular drug or drug class, search
all three drug code systems and drug_name which includes name and
dosage information combined. Below is an example retrieving
antidepressant prescriptions.

<br>

### Antidepressant prescriptions

#### Search by BNF code

```{r}
bnf <- bio_record(project_dir, record = "gp_scripts") %>%
    filter(str_detect(bnf_code, "^04\\.?03\\.?0[1-4]")) %>%
    collect()
```

<br>

#### Search by Read 2 code

Here I'm using `bio_code_primary_care` to read
the UKB-supplied primary care code lookup tables to find Read 2 code
equivalents of BNF antidepressant codes.

```{r}
read2_from_bnf <- bio_code_primary_care(project_dir, "read_v2_drugs_bnf") %>%
    filter(str_detect(bnf_code, "^04\\.?03\\.?0[1-4]")) %>%
    pull(read_code)

# Make a regular expression for search filter
read2_rgx <- read2_from_bnf %>%
    str_replace_all("\\.", "\\\\.") %>%
    str_c("^", ., collapse = "|") %>%
    regex()

read2 <- bio_record(project_dir, record = "gp_scripts") %>%
    filter(str_detect(read_2, read2_rgx)) %>%
    collect() %>%
    mutate(read_2 = str_replace(read_2, "00$", ""))
```

The `mutate` drops a trailing "00" on some read2 codes (a UKB artifact?).

<br>

__Note.__ [Fabbri et al. (2021) Genetic and clinical characteristics of
treatment-resistant depression using primary care records in two UK
cohorts](https://pubmed.ncbi.nlm.nih.gov/33753889/) use an augmented
list of read 2 codes, which includes the additional codes: du6z. du61.
gdez. gdey. gde2. gde1.

This list is in supplementary material table 1, available from the UKB
resources folder on Rosalind/CREATE. (See the
paper's [code for treatment-resistant depression phenotype derivation,
as well as antidepressant extraction](https://github.com/chiarafabbri/MDD_TRD_study/blob/master/scripts/extract_diagn_ADs_TRD_pheno.R))

```{r}
# -curated antidepressants - PMID33753889
cf_path <- file.path(
    dirname(project_dir),
    "resources/pmid33753889_medrxiv_supplementary_table1.xlsx"
)

excel_sheets(cf_path)
```

<span style="background-color:grey">
<pre>
[1] "readv2_mood_psych_sub" "readv3_mood_psych_sub" "ADs_read_2_codes"      "bnf_drug_names"        "dmd_drug_names"        "read2_drug_names"
</pre>
</span>

<br>

Chiara's augmented list of read 2 codes is in the sheet "ADs_read_2_codes".
```{r}
# CF Read 2 code list
cf_path <- file.path(
    dirname(project_dir),
    "resources/pmid33753889_medrxiv_supplementary_table1.xlsx"
)

read2_from_cf <- read_excel(cf_path, sheet = "ADs_read_2_codes") %>%
    pull(read2_code)

read2_rgx <- read2_from_cf %>%
    str_replace_all("\\.", "\\\\.") %>%
    str_c("^", ., collapse = "|") %>%
    regex()

read2 <- bio_record(project_dir, record = "gp_scripts") %>%
    filter(str_detect(read_2, read2_rgx)) %>%
    collect() %>%
    mutate(read_2 = str_replace(read_2, "00$", ""))
```

<br>

#### Search by dm+d code and/or drug name

You could also search for known dm+d codes in column dmd_code, and/or
known drug names in drug_name, to retrieve any prescriptions with
missing BNF and Read 2 codes. For example, Chiara's list of curated
antidepressant drug names is included in the ukbkings dataset
`drug_dmd_antidep`.


```{r}
drug_name_rgx <- str_c(drug_dmd_antidep$dmd_name) %>%
    str_c(collapse = "|") %>%
    regex(ignore_case = TRUE)

other <- bio_record(project_dir, record = "gp_scripts") %>%
    filter(bnf_code == "" & read_2 == "") %>%
    filter(str_detect(drug_name, drug_name_rgx)) %>%
    collect()
```

The `filter` on empty strings (missing values) in bnf_code and read_2
code searches among prescriptions not already captured by a bnf or
read 2 code.

<br>

Merge prescriptions

```{r}
px <- bind_rows(bnf, read2, other) %>%
    distinct() %>%
    as_tibble()
```

<br>


## Initial clean and format

Replace empty strings with `NA`, fix column types, and drop
prescriptions without an associated date (if appropriate).

```{r}
px <- px %>%
    mutate(
        across(.fns = ~ na_if(.x, "")),
        across(read_2:quantity, as.character),
        issue_date = as_date(issue_date, format = "%d/%m/%Y"),
    ) %>%
    filter(!is.na(issue_date))
```

<br>

The drug_name and quantity columns include a mix of incomplete
information (e.g., drug, dosage, form (i.e., capsule, tablet, etc),
count, company) and style (upper/lower case, integer/floating point,
same form expressed in many ways (e.g., tab, tablet, tablets,
tablet(s))). A random sample of 20 rows shows some the above

<span style="background-color:grey">
<pre>
# A tibble: 20 × 8
       eid data_provider issue_date read_2 bnf_code        dmd_code  drug_name                                        quantity             
     <int>         <int> <date>     <chr>  <chr>           <chr>     <chr>                                            <chr>                
 1 XXXXXXX             X XXXX-XX-XX NA     04.03.03.00.00  NA        Citalopram 20mg tablets                          28 tablets           
 2 XXXXXXX             X XXXX-XX-XX NA     04.03.04.00.00  NA        Venlafaxine 75mg modified-release capsules       56 capsule(s) - 75 mg
 3 XXXXXXX             X XXXX-XX-XX NA     0403010N0AAAEAE NA        IMIPRAMINE HYDROCHLORIDE TABLETS 25MG            60.000               
 4 XXXXXXX             X XXXX-XX-XX da91.  NA              NA        NA                                               NA                   
 5 XXXXXXX             X XXXX-XX-XX da93.  NA              NA        NA                                               NA                   
 6 XXXXXXX             X XXXX-XX-XX NA     040301          NA        Dosulepin Hydrochloride CAPS 25MG                168.000              
 7 XXXXXXX             X XXXX-XX-XX NA     04.03.03.00.00  NA        Prozac 20mg capsules (Eli Lilly and Company Ltd) 1 OP                 
 8 XXXXXXX             X XXXX-XX-XX NA     04.03.03.00.00  NA        Fluoxetine 20mg capsules                         60 capsule           
 9 XXXXXXX             X XXXX-XX-XX NA     04.03.03.00.00  NA        Fluoxetine 20mg capsules                         60 capsules          
10 XXXXXXX             X XXXX-XX-XX NA     04.03.04.00.00  NA        Venlafaxine 75mg modified-release capsules       84 capsule(s)        
11 XXXXXXX             X XXXX-XX-XX NA     04.03.04.00.00  NA        Mirtazapine 45mg tablets                         28 tablet            
12 XXXXXXX             X XXXX-XX-XX NA     04.03.03.00.00  NA        Sertraline 100mg tablets                         7 tablet             
13 XXXXXXX             X XXXX-XX-XX da41.  04030300        NA        Fluoxetine 20mg capsules                         30.000               
14 XXXXXXX             X XXXX-XX-XX d712.  NA              321746008 AMITRIPTYLINE HCl tabs 25mg                      56.000               
15 XXXXXXX             X XXXX-XX-XX NA     04.03.01.00.00  NA        Amitriptyline 25mg tablets                       168 tablet(s) - 25 mg
16 XXXXXXX             X XXXX-XX-XX daBz.  NA              NA        NA                                               NA                   
17 XXXXXXX             X XXXX-XX-XX NA     04.03.01.00.00  NA        Amitriptyline 10mg tablets                       28 tablets           
18 XXXXXXX             X XXXX-XX-XX NA     04.03.01.00.00  NA        Clomipramine 50mg capsules                       168 capsules - 50 mg 
19 XXXXXXX             X XXXX-XX-XX NA     04.03.03.00.00  NA        Sertraline 100mg tablets                         14 tablets           
20 XXXXXXX             X XXXX-XX-XX d79z.  NA              NA        NA                                               NA  
</pre>
</span>

<br>

In this subset of prescriptions, there are 1,655 and 4,830
unique entries for drug_name and quantity respectively.

```{r}
px %>%
    distinct(drug_name) %>%
    nrow()
px %>%
    distinct(quantity) %>%
    nrow()
```

<span style="background-color:grey">
<pre>
[1] 1655
[1] 4830
</pre>
</span>

<br>

A first pass tidy up of these two columns significantly reduces the
number of unique entries.
```{r}
tidy_rgx <- "[\\*\\-\\&\\'\\,]|\\(.*[\\)\\.{3}]|\\.{3}|\\.000$|\\[.*\\]|[\\[\\]]"
# \\/ part of dosage

px <- px %>%
    mutate(
        across(c(drug_name, quantity), ~ {
            tolower(.x) %>%
                str_replace_all(tidy_rgx, " ") %>%
                str_trim() %>%
                str_squish()
        })
    )

px %>%
    distinct(drug_name) %>%
    nrow()
px %>%
    distinct(quantity) %>%
    nrow()
```

<span style="background-color:grey">
<pre>
[1] 1116
[1] 3568
</pre>
</span>

<br>

## Additional tidying

This data will require further tidying, some of which will be specific
to the subset of prescriptions you collect. In the following, I've
created a series of regular expressions to capture alternatives for the
same descriptor, then added a new columns for drug, dose, and form.


```{r}
tab_rgx <- c("tablets", "tablet", "tabs", "tab", "soltab") %>%
    str_c("\\b", ., "\\b", collapse = "|") %>%
    regex()

cap_rgx <- c("capsules", "capsule", "caps", "cap", "ca", "cps", "capsules\\d+mg") %>%
    str_c("\\b", ., "\\b", collapse = "|") %>%
    regex()

sol_rgx <- c("solution", "soln") %>%
    str_c("\\b", ., "\\b", collapse = "|") %>%
    regex()

drp_rgx <- c("drop", "drops") %>%
    str_c("\\b", ., "\\b", collapse = "|") %>%
    regex()

sus_rgx <- c("suspension", "susp", "sus") %>%
    str_c("\\b", ., "\\b", collapse = "|") %>%
    regex()

syr_rgx <- c("syrup", "syrp", "syr") %>%
    str_c("\\b", ., "\\b", collapse = "|") %>%
    regex()

liq_rgx <- c("liquid", "liq") %>%
    str_c("\\b", ., "\\b", collapse = "|") %>%
    regex()

eli_rgx <- c("elixir", "eli") %>%
    str_c("\\b", ., "\\b", collapse = "|") %>%
    regex()

mix_rgx <- c("mixture", "mix") %>%
    str_c("\\b", ., "\\b", collapse = "|") %>%
    regex()

px <- px %>%
    mutate(
        drug = word(drug_name, 1),
        dose = str_extract(drug_name, "\\d*\\ *mg|\\d*\\ *microgram") %>% str_trim(),
        form = case_when(
            str_detect(drug_name, tab_rgx) |
                str_detect(quantity, tab_rgx) ~ "tablet",
            str_detect(drug_name, cap_rgx) |
                str_detect(quantity, cap_rgx) ~ "capsule",
            str_detect(drug_name, sol_rgx) |
                str_detect(quantity, sol_rgx) ~ "solution",
            str_detect(drug_name, drp_rgx) |
                str_detect(quantity, drp_rgx) ~ "drops",
            str_detect(drug_name, sus_rgx) |
                str_detect(quantity, sus_rgx) ~ "suspension",
            str_detect(drug_name, syr_rgx) |
                str_detect(quantity, syr_rgx) ~ "syrup",
            str_detect(drug_name, liq_rgx) |
                str_detect(quantity, liq_rgx) ~ "liquid",
            str_detect(drug_name, eli_rgx) |
                str_detect(quantity, eli_rgx) ~ "elixir",
            str_detect(drug_name, mix_rgx) |
                str_detect(quantity, mix_rgx) ~ "mixture"
        )
    )
```


This is not an exhautive clean of the data. You could continue to clean
this data by iteratively running the above chunk, then inspecting the
remaining unique entries in drug_name for missing values in drug/ form/
dose as follows

```{r}
px %>%
    filter(is.na(form)) %>%
    distinct(drug_name) %>%
    print(n = Inf)
```


Add further terms to the regular expressions above, and repeat until the
data are sufficiently clean for your purposes.

<br>

## 3. Other medication reference data

|data|source|
|---|---|
| drug_pharmgkb | PharmGKB curated drug list (​https://www.pharmgkb.org/downloads​, drugs.zip) |
| drug_gwas | ukb.gwas.medication.supplementary.data.1.xlsx - Genome-wide association study of medication-use and associated disease in the UK Biobank, Nature Communications |
| drug_dmd_antidep | A curated list of dm+d antidepressant drug names from Fabbri et al. |


## 4. Resources

__NHS prescription drugs__
- https://openprescribing.net/bnf/
- https://www.thedatalab.org/blog/161/prescribing-data-bnf-codes/
- https://digital.nhs.uk/services/terminology-and-classifications/read-codes
- https://digital.nhs.uk/data-and-information/areas-of-interest/prescribing/practice-level-prescribing-in-england-a-summary

__ATC codes__
- https://ec.europa.eu/health/documents/community-register/html/reg_hum_atc.htm
- https://bioportal.bioontology.org/ontologies/ATC/?p=summary