---
title: "Using UKB primary care prescription data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using UKB primary care prescription data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    eval = FALSE,
    collapse = TRUE,
    comment = "#>"
)
```

```{r setup}
library(ukbkings)
```

## Query prescription data

UKB medication presciption data are included in the
[Primary Care Linked data](https://biobank.ndph.ox.ac.uk/showcase/showcase/docs/primary_care_data.pdf),
accessed through the UKB Data Portal. The data can be
explored online with SQL queries, the results of which can be
downloaded, or the full table can be downloaded. We have downloaded
the full table and converted it to an R disk.frame using ukbproject for
fast local queries with ukbkings `bio_record` and `bio_record_map`.

`bio_record` returns a diskframe, an object which can be queried without
reading the data into R.

```{r}
project_dir <- "<path_to_project_directory>"

scripts_diskf <- bio_record(project_dir, record = "gp_scripts")

glimpse(scripts_diskf)
```


Apply dplyr verbs to subset rows and columns as required and `collect`
to read the data into R as a dataframe.

```{r filter}
scripts <- gp_scripts_diskf %>%
    filter(str_detect(drug_name, "fluoxetine")) %>%
    select(eid, data_provider, issue_date) %>%
    collect()
```


To extract data for a only a subset of samples use the `subset`
argument to `bio_record` which will now return a dataframe. __Note__.
You do not need to `collect` when using the `subset` argument.
(Additional row and column filtering can be applied as above)

```{r sample subset}
sample_subset <- c(321, 654, 987)

scripts <- bio_record(
    project_dir,
    record = "gp_scripts",
    subset = sample_subset
)
```


***


## Retrieve antidepressant prescriptions

Described below is a ukbkings implementation of the extraction performed
by [Fabbri et al. (2021) Genetic and clinical characteristics of
treatment-resistant depression using primary care records in two UK
cohorts](). (See the original paper's [code for treatment-resistant
depression phenotype derivation, as well as antidepressant
extraction](https://github.com/chiarafabbri/MDD_TRD_study/blob/master/scripts/extract_diagn_ADs_TRD_pheno.R))


To extract all antidepressant prescriptions

```{r antidepressant extraction}
antidep_prescriptions <- bio_med(project_dir, chapter = "antidepressants")
```


This is equivalent to

```{r}
# Antidepressant prescriptions

# Replace with one filter with bnf OR dmd OR read2 code, or

# Search by BNF code
bnf_antidep_prescriptions <- bio_record(project_dir, record = "gp_scripts") %>%
    filter(str_detect(bnf_code, "^04\\.?03\\.?0[1-4]")) %>%
    collect() %>%
    distinct()

# Search by dm+d code
dmd_antidep_prescriptions <- bio_record(project_dir, record = "gp_scripts") %>%
    filter()

# Search by Read 2 code
read2_antidep_prescriptions <- bio_record(project_dir, record = "gp_scripts") %>%
    filter()
```


### Medication codes

| code | meaning |
|---|---|
| bnf_code | British National Formulary (BNF) is a reference book containing the standard list of medicines used in UK |
| dmd_code | Dictionary of Medicines and Devices (dm+d), the catalogue of things that can be prescribed in the NHS |
| read_2 | Read Codes (version 2 (v2)) are a coded thesaurus of clinical terms (used in the NHS since 1985). There is also a version 3 (CTV3 or v3)  |


#### BNF code

UKB captures only the first few pieces of BNF code information:
Chapter.Section.Paragraph.Sub-paragraph.Chemical substance, e.g.,
04.03.01 tricyclic and related antidepressants,
04.03.02 monoamine-oxidase inhibitors, etc.

__BNF Chapter 4: Central Nervous System__

- 4.1: Hypnotics and anxiolytics
- 4.2: Drugs used in psychoses and related disorders
- __4.3: Antidepressant drugs__
    - __4.3.1: Tricyclic and related antidepressant drugs__
    - __4.3.2: Monoamine-oxidase inhibitors (maois)__
    - __4.3.3: Selective serotonin re-uptake inhibitors__
    - __4.3.4: Other antidepressant drugs__
- 4.4: CNS stimulants and drugs used for ADHD
- 4.5: Drugs used in the treatment of obesity
- 4.6: Drugs used in nausea and vertigo
- 4.7: Analgesics
- 4.8: Antiepileptic drugs
- 4.9: Drugs used in parkinsonism and related disorders
- 4.10: Drugs used in substance dependence
- 4.11: Drugs for dementia


### Medication dose


## Other medication reference data

|data|source|
|---|---|
| meds_pharmgkb | PharmGKB curated drug list (​https://www.pharmgkb.org/downloads​, drugs.zip) |
| meds_gwas | ukb.gwas.medication.supplementary.data.1.xlsx - Genome-wide association study of medication-use and associated disease in the UK Biobank, Nature Communications |
| meds_dmd_antidep | A curated list of dm+d antidepressant drug names from Fabbri et al. |
| meds_ukb | Nested dataframe of UKB primary care drug coding maps |


## Resources

- https://openprescribing.net/bnf/
- https://www.thedatalab.org/blog/161/prescribing-data-bnf-codes/
- https://digital.nhs.uk/services/terminology-and-classifications/read-codes
- https://digital.nhs.uk/data-and-information/areas-of-interest/prescribing/practice-level-prescribing-in-england-a-summary
